### 2024-11-27 paper/paper.qmd
start writing on Data and Model sections

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Kevin Cai
thanks: "Code and data are available at: [https://github.com/kevicai/toronto-healthcare-outbreak-prediction](https://github.com/kevicai/toronto-healthcare-outbreak-prediction)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
```

# Introduction {#sec-intro}

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

This report uses the Outbreaks in Toronto Healthcare Institutions dataset, contains data from January 2016 to November 2024. The dataset is provided by Toronto Public Health, through City of Toronto Open Data Portal [@datasource]. The dataset tracks reported outbreaks of gastroenteric and respiratory illnesses in Toronto healthcare institutions and contains detailed information on outbreak settings, causative agents, and outbreak durations. Following the principles from Telling Stories with Data [@textbook], we examine how the characteristics of outbreaks, such as the type of healthcare institution, the causative agent, and the month the outbreak began, influence their durations. A sample of the cleaned dataset is shown in @tbl-cleaned.

```{r, message = FALSE, echo = FALSE}
#| label: tbl-cleaned
#| tbl-cap: "Sample of Cleaned Outbreaks in Toronto Healthcare Institution Data"
data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
head(data, 5) |>
    kable(col.names = c("Outbreak Setting", "Causative Agent", "Month", "Outbreak Duration"))
```

There is 5387 observations in the orginal dataset and 1119 observations were removed that contained missing, invalid, or irrelivant data of the variables we're interested in. The data was first downloaded using `Python` [@citePython] and cleaned with the `pandas` package [@citePandas]. The cleaning process involved converting dates to a standardized datetime format, creating a "duration" variable representing the length of each outbreak, and extracting the month of the outbreak's start. Irrelevant columns were removed, and variables were renamed for clarity. Causative agents were grouped into broader categories, and rows with missing or invalid data were removed, including those with unidentifiable causative agents or certain outbreak settings. The final dataset was saved for further analysis.

`R` [@citeR] is used for the generation of figures, graphs, and tables throughout this paper. Specifically, the `rstanarm` package [@rstanarm] was employed to fit the model. For data manipulation, the `dplyr` package [@dplyr] was utilized to clean and transform the data efficiently. The `caret` package [@caret] was used for model training, while `modelsummary` [@modelsummary] was used to produce concise tables summarizing the model output. The `loo` package [@loo] was used to perform leave-one-out cross-validation, which helped assess the model’s predictive performance.

## Measurement

The data was primarily collected through mandatory reporting by healthcare institutions to Toronto Public Health under the Ontario Health Protection and Promotion Act (HPPA). Reports of suspected or confirmed outbreaks include both gastroenteric and respiratory illnesses. These reports are based on active monitoring by institutional staff, who observe and document signs and symptoms such as nausea, vomiting, fever, cough, or sore throat.

Some details, such as the causative agent group, may initially be unconfirmed and later identified through laboratory tests or clinical evaluations. However, these identifications are not always definitive. For instance, "Coronavirus*" in the dataset refers to seasonal coronaviruses, which are commonly implicated in respiratory outbreaks, and does not include COVID-19.

The unit of measurement for outbreak duration is in days. Other data fields, such as outbreak setting and causative agent group, are categorical features without numerical units. The dataset is updated weekly, ensuring it reflects the most recent outbreak data available.

## Outcome variable

### Duration
The Duration variable is numerical and indicates the total number of days each outbreak lasted. This reflects the severity and magnitude of the outbreak. It is constructed from the dataset by calculating the difference between the outbreak start and end dates.

```{r, message = FALSE, echo = FALSE}
#| label: fig-duration
#| fig-cap: "Distribution of Outbreak Duration"
ggplot(data, aes(x = duration)) +
  geom_histogram(fill = "steelblue", binwidth = 5, alpha = 0.8) +
  labs(
    x = "Duration (Days)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10)
  )
```

Longer outbreak durations may indicate challenges in containment, possibly influenced by the Outbreak Setting and Causative Agent.

## Predictor variables

### Outbreak Setting
The Outbreak Setting variable is categorical and identifies the type of healthcare institution where the outbreak occurred, such as hospitals, long-term care homes (LTCH), or retirement homes. It provides insights into the environments most affected by outbreaks.

@fig-setting-count illustrates the count of outbreaks across different settings in the dataset.

```{r, message = FALSE, echo = FALSE}
#| label: fig-setting-count
#| fig-cap: "Outbreak occurrence in healthcare settings"
outbreak_setting_count <- data %>%
  count(outbreak_setting, name = "count") %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(count))

# Bar graph for Outbreak Setting
ggplot(outbreak_setting_count, aes(x = reorder(outbreak_setting, -count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.1, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.2))) + # Add right padding
  coord_flip() + 
  labs(
    x = "Outbreak Setting",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10)
  )
```

LTCH (Long-Term Care Homes) accounts for a significant portion of outbreaks, likely due to the vulnerability of their populations. Comparing the frequency of outbreaks across settings can reveal risk patterns.

### Causative Agent
The Causative Agent variable is categorical and reflects the infectious agents responsible for outbreaks. While the original dataset contains 55 agents, they are grouped into seven broader categories to simplify the analysis and enhance interpretability.

@fig-agent-count illustrates the count and percentage distribution of causative agents in the dataset.

```{r, message = FALSE, echo = FALSE}
#| label: fig-agent-count
#| fig-cap: "Outbreak causative agent count and percentage"
outbreak_agent_count <- data %>%
  count(causative_agent, name = "count") %>%
  mutate(percentage = (count / sum(count)) * 100) %>%
  arrange(desc(count))

# Bar graph for Causative Agents
ggplot(outbreak_agent_count, aes(x = reorder(causative_agent, -count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.1, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.2))) + # Add right padding
  coord_flip() +
  labs(
    x = "Causative Agent",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.margin = margin(1, 1.2, 1, 1, "cm")
  )
```

### Month Outbreak Began
The Month variable is numerical and records the calendar month when each outbreak started. It reflects seasonal trends and potential patterns in infection rates. This variable is extracted from the date where each outbreak began from the original dataset.

@fig-month shows the the occurance of outbreaks in each month, with winter months having siginifiantly more outbreaks compared to other months. This suggest that seasons have effects on outbreak occurances.

@fig-month-duration the boxplot visualizes the distribution of outbreak durations for each month. While winter months have a higher frequency of outbreaks, the duration of these outbreaks appears similar across all months, with no distinct seasonal differences in median or variability of duration.

```{r, message = FALSE, echo = FALSE}
#| label: fig-month
#| fig-cap: "Seasonal trends in outbreak occurrence and percentage"
outbreak_month_count <- data %>%
  count(month, name = "count") %>%
  arrange(desc(month))

ggplot(outbreak_month_count, aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    x = "Month",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10)
  )
```

```{r, message = FALSE, echo = FALSE}
#| label: fig-month-duration
#| fig-cap: "Duration of outbreaks across different months"
library(ggplot2)

# Scatter plot for month vs. duration
ggplot(data, aes(x = as.factor(month), y = duration)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(
    x = "Month",
    y = "Duration (days)",
    title = "Duration of Outbreaks Across Months"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  )
```

# Model

## Model Overview

To better understand the factors influencing the duration of outbreaks in Toronto healthcare facilities, a statistical model was developed using the negative binomial regression framework. This model was chosen because the outcome variable of interest, outbreak duration, is a count variable with evidence of overdispersion—where the variance exceeds the mean [@textbook]. Additionally, this model was Bayesian, meaning the parameters were treated as random variables with prior probability distributions reflecting initial beliefs about their values before considering the data.

Both models were implemented  using the `rstanarm` package.

## Model Setup

The setup for the Bayesian Negative Binomial regression model used in this analysis is as follows:

\begin{align} 
y_i | \lambda_i &\sim \text{Negative Binomial}(\lambda_i, \phi) \\
\log(\lambda_i) &= \beta_0 + \beta_1 \times \text{outbreak\_setting}_i + \beta_2 \times \text{causative\_agent}_i + \beta_3 \times \text{month}_i \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\phi &\sim \text{Exponential}(1)
\end{align}

In the above model:

- $\mu_i$ is the predicted duration of the outbreak $i$ in days.
- $\lambda_i$ is the expected duration of outbreak $i$, modeled through a **log link**.
- $\beta_0$ is the intercept term.
- $\beta_1$ is the coefficient for the **outbreak setting**.
- $\beta_2$ is the coefficient for the **causative agent**.
- $\beta_3$ is the coefficient for the **month** when the outbreak started.
- $\phi$ is the **dispersion parameter** that controls the degree of overdispersion in the Negative Binomial distribution.

- All coefficients ($\beta_0, \beta_1, \beta_2, \beta_3$) are assigned **Normal(0, 2.5)** priors.
- The dispersion parameter $\phi$ is assigned an **Exponential(1)** prior, reflecting a non-informative prior belief about the variance.

## Model Selection

```{r, message = FALSE, echo = FALSE}
#| label: load-models
library(rstanarm)#|

poisson_model <-
    readRDS(file = here::here("models/poisson_model.rds"))
neg_binom_model <- 
    readRDS(file = here::here("models/neg_binomial_model.rds"))
```

Both Negative Binomial model and Poisson model was constructed using rstanarm for the dataset. Negative Binomial model was chosen over the Poisson model as overall it was better at effectively predicting the  because it performed better in terms of the expected log pointwise predictive density (ELPD). We found that the Negative Binomial model has a larger ELPD, indicating that it provides a better fit to the data.

## Model Diagnostics and Validation

We performed several model validation checks to assess the adequacy of both models.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

# modelsummary::modelsummary(
#     list(
#         "Model" = model
#     ),
#     statistic = "mad",
#     fmt = 2
# )
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Outcome Variable Variance and Mean

## Posterior predictive check

In @fig-ppchecks, using code adapted from @textbook, posterior prediction checks were performed for both the Poisson model and the Negative Binomial model. The figure show how well the model is able to predict the observed outcomes.

```{r, message = FALSE, echo = FALSE}
#| label: fig-ppchecks
#| layout-ncol: 2
#| fig-cap: "Comparing posterior prediction checks for the Poisson model and the Negative Binomial model"
#| fig-subcap: ["Poisson model posterior prediction check", "Negative Binomial model posterior prediction check"]
pp_check(poisson_model) +
  theme_classic() +
  theme(legend.position = "bottom")

pp_check(neg_binom_model) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Diagnostics

<!-- ```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(model, "trace")

plot(model, "rhat")
``` -->

\newpage

# References

### 2024-11-27 scripts/00-simulate_data.py

@@ -21,10 +21,9 @@ np.random.seed(853)
 outbreak_settings = [
     "Hospital-Chronic Care",
     "Hospital-Psychiatric",
+    "Hospital−Acute Care",
     "LTCH",
     "Retirement Home",
-    "Shelter",
-    "Transitional Care",
 ]
 
 # Causative agent groups

### 2024-11-27 scripts/03-clean_data.py

@@ -119,7 +119,7 @@ rows_before = df.shape[0]
 # Filter invalid or uninformative rows
 df = df.dropna()
 df = df[df["Main Causative Agent"] != "Unable to identify"]
-df = df[~df["outbreak_setting"].isin(["Shelter", "Transitional care"])]
+df = df[~df["outbreak_setting"].isin(["Shelter", "Transitional Care"])]
 
 def group_causative_agent(agent):
     mapping = reverse_mapping.get(agent, "Other")

### 2024-11-27 scripts/07-model_data.R

@@ -33,12 +33,13 @@ X <- train_data %>%
   select(outbreak_setting, causative_agent, month)
 y <- train_data$duration
 
-# Fit Poisson and Negative Binomial models 
+# Fit Poisson and Negative Binomial models
 poisson_model <- stan_glm(
   duration ~ outbreak_setting + causative_agent + month,
   data = train_data,
   family = poisson(link = "log"),
-  prior = normal(0, 2.5),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
   seed = 22
 )
 
@@ -46,7 +47,8 @@ neg_binomial_model <- stan_glm(
   duration ~ outbreak_setting + causative_agent + month,
   data = train_data,
   family = neg_binomial_2(link = "log"),
-  prior = normal(0, 2.5),
+  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
   seed = 22
 )
 

### 2024-11-27 scripts/08-check_models.R

@@ -10,7 +10,7 @@
 
 # Load the previously saved models
 poisson_model <- readRDS("models/poisson_model.rds")
-neg_binomial_model <- readRDS("models/neg_binom_model.rds")
+neg_binomial_model <- readRDS("models/neg_binomial_model.rds")
 
 # Load test data from CSV
 test_data <- read.csv("data/02-analysis_data/test_data.csv")
@@ -21,13 +21,12 @@ test_data <- test_data %>%
     month = factor(month)
   )
 
-
 # Posterior predictive checks
-# pp_check(poisson_model) +
-#   theme(legend.position = "bottom")
+pp_check(poisson_model) +
+  theme(legend.position = "bottom")
 
-# pp_check(neg_binomial_model) +
-#   theme(legend.position = "bottom")
+pp_check(neg_binomial_model) +
+  theme(legend.position = "bottom")
 
 # Compare the models using LOO (Leave-One-Out Cross Validation)
 poisson_loo <- loo(poisson_model, cores = 2)

### 2024-11-27 paper/paper.qmd
add title

@@
 ---
-title: "My title"
+title: "Understanding the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
 subtitle: "My subtitle if needed"

### 2024-11-27 paper/paper.qmd
complete Model section of paper

@@
 library(knitr)
+library(ggplot2)
+library("rstanarm")
+library("caret")
+library("loo")
 ```
@@
 The `loo` package [@loo] was used to perform leave-one-out cross-validation, which helped assess the model’s predictive performance.
+Finally, the package `ggplot2` is used to generate graphics and figures for this analysis.
 
@@
 
+```{r, message = FALSE, echo = FALSE}
+#| label: tbl-duration-mean-var
+#| tbl-cap: "Summary of Outbreak Duration: Mean and Variance"
+mean_duration <- mean(data$duration)
+var_duration <- var(data$duration)
+
+tibble(
+  Statistic = c("Mean Duration", "Variance"),
+  Value = c(mean_duration, var_duration)
+) |> 
+  kable(col.names = c("Statistic", "Value"))
+```
+
 Longer outbreak durations may indicate challenges in containment, possibly influenced by the Outbreak Setting and Causative Agent.
@@
 #| fig-cap: "Outbreak occurrence in healthcare settings"
-outbreak_setting_count <- data %>%
-  count(outbreak_setting, name = "count") %>%
-  mutate(percentage = (count / sum(count)) * 100) %>%
+outbreak_setting_count <- data |>
+  count(outbreak_setting, name = "count") |>
+  mutate(percentage = (count / sum(count)) * 100) |>
   arrange(desc(count))
@@
 #| fig-cap: "Outbreak causative agent count and percentage"
-outbreak_agent_count <- data %>%
-  count(causative_agent, name = "count") %>%
-  mutate(percentage = (count / sum(count)) * 100) %>%
+outbreak_agent_count <- data |>
+  count(causative_agent, name = "count") |>
+  mutate(percentage = (count / sum(count)) * 100) |>
   arrange(desc(count))
@@
 @fig-month-duration the boxplot visualizes the distribution of outbreak durations for each month.
-While winter months have a higher frequency of outbreaks, the duration of these outbreaks appears similar across all months, with no distinct seasonal differences in median or variability of duration.
+The duration of months January to November outbreaks appears similar, while December has a noticeable increase in duration compared to other months.
 
@@
 #| fig-cap: "Seasonal trends in outbreak occurrence and percentage"
-outbreak_month_count <- data %>%
-  count(month, name = "count") %>%
+outbreak_month_count <- data |>
+  count(month, name = "count") |>
   arrange(desc(month))
@@
 #| fig-cap: "Duration of outbreaks across different months"
-library(ggplot2)
-
 # Scatter plot for month vs. duration
@@
 
-Both models were implemented  using the `rstanarm` package.
-
 ## Model Setup
 
-The setup for the Bayesian Negative Binomial regression model used in this analysis is as follows:
+The setup for the Bayesian negative binomial regression model used in this analysis is as follows:
 
@@
 
-- $\mu_i$ is the predicted duration of the outbreak $i$ in days.
-- $\lambda_i$ is the expected duration of outbreak $i$, modeled through a **log link**.
+- $\lambda_i$ is the expected duration of outbreak $i$, modeled through a log link.
 - $\beta_0$ is the intercept term.
@@
 - $\beta_3$ is the coefficient for the **month** when the outbreak started.
-- $\phi$ is the **dispersion parameter** that controls the degree of overdispersion in the Negative Binomial distribution.
+- $\phi$ is the **dispersion parameter** that controls the degree of overdispersion in the negative binomial distribution.
 
 - All coefficients ($\beta_0, \beta_1, \beta_2, \beta_3$) are assigned **Normal(0, 2.5)** priors.
-- The dispersion parameter $\phi$ is assigned an **Exponential(1)** prior, reflecting a non-informative prior belief about the variance.
+- The dispersion parameter $\phi$ is assigned an **Exponential(1)** prior, reflecting our belief that the variance is greater than the mean.
 
@@
 #| label: load-models
-library(rstanarm)#|
-
+library(rstanarm)
 poisson_model <-
     readRDS(file = here::here("models/poisson_model.rds"))
-neg_binom_model <- 
+neg_binomial_model <- 
     readRDS(file = here::here("models/neg_binomial_model.rds"))
@@
 
-Both Negative Binomial model and Poisson model was constructed using rstanarm for the dataset.
-Negative Binomial model was chosen over the Poisson model as overall it was better at effectively predicting the  because it performed better in terms of the expected log pointwise predictive density (ELPD).
-We found that the Negative Binomial model has a larger ELPD, indicating that it provides a better fit to the data.
+Both negative binomial model and Poisson model for the dataset was constructed using the `rstanarm` package [@rstanarm] and R [@citeR].
+But the negative binomial model was chosen over the Poisson model for several reasons.
+First, as shown in @tbl-modelresults, the variance of the outcome variable, duration, is significantly higher than the mean, indicating overdispersion.
+The Poisson model assumes equal mean and variance, which is not suitable in this case.
+The negative binomial model relaxes this assumption, allowing for overdispersion and providing a better fit for the data [@textbook].Additionally, the Leave-One-Out Cross Validation (LOO-CV) results in @tbl-loo-comparison show that the negative binomial model has a higher ELPD (Expected Log Pointwise Predictive Density) compared to the Poisson model.
+The ELPD is a metric that measures the model’s predictive performance, with higher values indicating a better fit to the data  [@textbook].
+The fact that the negative binomial model outperforms the Poisson model in this regard suggests that it is more effective at capturing the underlying patterns of the outbreak duration data.
+
+Other regression models like logistic regression were not chosen because logistic regression is designed for modeling binary outcomes.
+Since our outcome variable, duration, is a continuous count variable representing the number of days an outbreak lasts, logistic regression is not appropriate because it cannot model continuous or count data.
+Linear regression was also not chosen because Poission and negative binomial distributions are more suitable for modeling count data like outbreak duration in days, where as linear regression is more suitible for continuous data.
 
@@
 
-We performed several model validation checks to assess the adequacy of both models.
+We conducted several key validation checks to assess its predictive performance and overall adequacy.
+Aside from using LOO Cross Validation technique, we also calculated the Mean Absolute Error (MAE) for both models as a metric to assess the predictive performance of the Negative Binomial model over the Poisson model.
+To ensure the model doesn't over fit the training data, we first split the data into training and test sets.
+The data was randomly divided using the `caret` package [@caret], with 80% used for model training and the remaining 20% reserved for testing.
+We used both models to predict the outcome variable (outbreak duration) on the test set and compared the predicted values to the actual values from the test set to compute the MAE for each model.
+
+```{r, message = FALSE, echo = FALSE}
+#| label: tbl-mae-comparison
+#| tbl-cap: "Comparison of Mean Absolute Error (MAE) for Poisson and Negative Binomial Models"
+test_data <- read.csv(here::here("data/02-analysis_data/test_data.csv"))
+
+# Convert variables to factors
+test_data <- test_data |>
+  mutate(
+    outbreak_setting = factor(outbreak_setting),
+    causative_agent = factor(causative_agent),
+    month = factor(month)
+  )
+
+poisson_pred <- predict(poisson_model, newdata = test_data, type = "response")
+neg_binomial_pred <- predict(neg_binomial_model, newdata = test_data, type = "response")
+
+# Compute MAEs and present in a table
+mae_poisson <- round(mean(abs(test_data$duration - poisson_pred)), 2)
+mae_neg_binomial <- round(mean(abs(test_data$duration - neg_binomial_pred)), 2)
+
+mae_table <- tibble(
+  Model = c("Poisson Model", "Negative Binomial Model"),
+  MAE = c(mae_poisson, mae_neg_binomial)
+)
+
+mae_table |>
+  kable(col.names = c("Model", "Mean Absolute Error (MAE)")) 
+```
+
+From @tbl-mae-comparison, the MAE for the Poisson model is 6.53, while the MAE for the Negative Binomial model is 6.52.
+The difference between the MAEs is minimal, suggesting that both models perform similarly in terms of prediction accuracy.
+However, the Negative Binomial model may still be preferred as it accounts for overdispersion, which is more appropriate for count data.
+The MAE of 6.52 means that, on average, the predicted outbreak duration from the Negative Binomial model deviates from the actual duration by approximately 6.52 days.
+In other words, for any given outbreak in the test data, the model's prediction of the outbreak's duration is off by around 6.5 days, either overestimating or underestimating the actual duration.
 
@@
 
-Our results are summarized in @tbl-modelresults.
+Our results are summarized in @tbl-model-results.
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
+```{r, message = FALSE, echo = FALSE}
+#| label: tbl-model-results
 #| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
 
@@
 
-In @fig-ppchecks, using code adapted from @textbook, posterior prediction checks were performed for both the Poisson model and the Negative Binomial model.
+In @fig-ppchecks, using code adapted from @textbook, posterior prediction checks were performed for both the Poisson model and the negative binomial model.
 The figure show how well the model is able to predict the observed outcomes.
@@
 #| layout-ncol: 2
-#| fig-cap: "Comparing posterior prediction checks for the Poisson model and the Negative Binomial model"
-#| fig-subcap: ["Poisson model posterior prediction check", "Negative Binomial model posterior prediction check"]
+#| fig-cap: "Comparing posterior prediction checks for the Poisson model and the negative binomial model"
+#| fig-subcap: ["Poisson model posterior prediction check", "Negative binomial model posterior prediction check"]
 pp_check(poisson_model) +
@@
 
-pp_check(neg_binom_model) +
+pp_check(neg_binomial_model) +
   theme_classic() +
   theme(legend.position = "bottom")
+```
+
+## Leave-One-Out (LOO) Cross Validation (CV) Comparison
+
+In @fig-loo-comparison, we compare LOO performance of the Poission model against the negative binomial model based on the expected log pointwise predictive density (ELPD) and find that the negative binomial model has a higher ELPD value.
+
+```{r, message = FALSE, echo = FALSE}
+#| label: tbl-loo-comparison
+#| tbl-cap: "Comparing LOO for Poisson and negative binomial models"
+poisson_loo <- loo(poisson_model, cores = 2)
+neg_binomial_loo <- loo(neg_binomial_model, cores = 2)
+
+loo_comparison <- loo_compare(poisson_loo, neg_binomial_loo)
+loo_comparison
 ```

### 2024-11-27 scripts/07-model_data.R

@@ -20,8 +20,8 @@ set.seed(22)
 # Load the previously saved training and test data
 train_data <- read.csv("./data/02-analysis_data/train_data.csv")
 
-# Convert categorical variables to factors
-train_data <- train_data %>%
+# Convert variables to factors
+train_data <- train_data |>
   mutate(
     outbreak_setting = factor(outbreak_setting),
     causative_agent = factor(causative_agent),
@@ -29,7 +29,7 @@ train_data <- train_data %>%
   )
 
 # Separate the features and target variable for modeling
-X <- train_data %>%
+X <- train_data |>
   select(outbreak_setting, causative_agent, month)
 y <- train_data$duration
 

### 2024-11-27 scripts/08-check_models.R

@@ -8,37 +8,39 @@
 # - 07-model_data.R
 # - Install required R packages: 'rstanarm', 'dplyr', 'caret', 'loo'
 
+library("dplyr")
+library("rstanarm")
+library("caret")
+library("loo")
+
 # Load the previously saved models
 poisson_model <- readRDS("models/poisson_model.rds")
 neg_binomial_model <- readRDS("models/neg_binomial_model.rds")
 
 # Load test data from CSV
 test_data <- read.csv("data/02-analysis_data/test_data.csv")
-test_data <- test_data %>%
+
+# # Convert variables to factors
+test_data <- test_data |>
   mutate(
     outbreak_setting = factor(outbreak_setting),
     causative_agent = factor(causative_agent),
     month = factor(month)
   )
 
-# Posterior predictive checks
-pp_check(poisson_model) +
-  theme(legend.position = "bottom")
+# # Posterior predictive checks
+# pp_check(poisson_model) +
+#   theme(legend.position = "bottom")
 
-pp_check(neg_binomial_model) +
-  theme(legend.position = "bottom")
+# pp_check(neg_binomial_model) +
+#   theme(legend.position = "bottom")
 
-# Compare the models using LOO (Leave-One-Out Cross Validation)
-poisson_loo <- loo(poisson_model, cores = 2)
-neg_binomial_loo <- loo(neg_binomial_model, cores = 2)
+# # Compare the models using LOO (Leave-One-Out Cross Validation)
+# poisson_loo <- loo(poisson_model, cores = 2)
+# neg_binomial_loo <- loo(neg_binomial_model, cores = 2)
 
-loo_comparison <- loo_compare(poisson_loo, neg_binomial_loo)
-print(loo_comparison)
-
-# Summarize the results of the LOO comparison
-cat("Summary of LOO Comparison:\n")
-cat("ELPD Difference:", loo_comparison$elpd_diff, "\n")
-cat("SE Difference:", loo_comparison$se_diff, "\n")
+# loo_comparison <- loo_compare(poisson_loo, neg_binomial_loo)
+# loo_comparison
 
 # Make predictions on the test data
 poisson_pred <- predict(poisson_model, newdata = test_data, type = "response")
@@ -49,7 +51,3 @@ neg_binomial_pred <- predict(neg_binomial_model, newdata = test_data, type = "re
 mae_poisson <- mean(abs(test_data$duration - poisson_pred))
 
[…]

### 2024-11-27 paper/paper.qmd
update readme

@@
 ---
-title: "Understanding the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
+title: "Analyzing the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
 subtitle: "My subtitle if needed"

### 2024-11-28 paper/paper.qmd
add data files

@@
 number-sections: true
+execute: 
+  cache: true
 bibliography: references.bib
@@
 library(ggplot2)
-library("rstanarm")
-library("caret")
-library("loo")
+library(rstanarm)
+library(caret)
+library(loo)
+library(broom.mixed)
+library(modelsummary)
+library(kableExtra)
 ```
@@
 
-Overview paragraph
+Outbreaks of infectious diseases in healthcare institutions are a significant challenge, particularly in hospitals, long-term care homes, and retirement homes, where vulnerable populations are most at risk.
+Understanding the factors that influence the duration of these outbreaks is important for improving preparedness, response times, and resource allocation.
+Previous studies have looked at factors affecting outbreak durations, but more research is needed, especially in Toronto’s healthcare system.
+This paper analyzes the key factors that influence the duration of outbreaks in Toronto healthcare institutions, focusing on outbreak setting, causative agents, and the timing of outbreaks.
 
-Estimand paragraph
+This study uses a Bayesian negative binomial regression model to predict outbreak durations in Toronto's healthcare facilities.
+The model includes factors like the type of healthcare institution (outbreak setting), the causative agent, and the month the outbreak occurred.
+This research aims to identify which factors contribute most to variation in outbreak durations.
+It provides actionable information for public health officials and healthcare managers to improve their response strategies.
 
-Results paragraph
+The results of this analysis help manage outbreaks, optimize patient care, and reduce the strain on healthcare facilities.
+Identifying high-risk settings or times of year when outbreaks last longer can improve resource allocation and response planning.
+This study also adds to the literature on healthcare outbreaks, with a focus on Toronto’s healthcare system.
 
-Why it matters paragraph
-
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+The remainder of this paper is structured as follows: @sec-data covers the dataset used for analysis, @sec-model describes the model setup and methodology, @sec-results presents the results and their interpretation, and @sec-discussion discusses the findings and future research directions.
 
@@
 Finally, the package `ggplot2` is used to generate graphics and figures for this analysis.
+The starter code and the data analysis techniques used are from Telling Stories with Data (@textbook).
 
@@
 ### Month Outbreak Began
-The Month variable is numerical and records the calendar month when each outbreak started.
+The Month variable is categorical and records the calendar month when each outbreak started as a name (e.g., Jan, Feb).
 It reflects seasonal trends and potential patterns in infection rates.
-This variable is extracted from the date where each outbreak began from the original dataset.
+This variable is extracted from the date where each outbreak began from the original dataset and converted from a number to the corresponding month name.
 
@@
 
-- All coefficients ($\beta_0, \beta_1, \beta_2, \beta_3$) are assigned **Normal(0, 2.5)** priors.
-- The dispersion parameter $\phi$ is assigned an **Exponential(1)** prior, reflecting our belief that the variance is greater than the mean.
+All coefficients ($\beta_0, \beta_1, \beta_2, \beta_3$) are assigned the prior of Normal(0, 2.5), which is the `rstanarm` package's default priors.
+The choice of  the prior is sufficient for the model in this analysis and is a common non-informative prior that allows for reasonable variability [@textbook].
+The dispersion parameter $\phi$ is assigned an **Exponential(1)** prior, which reflects the observation that the variance is greater than the mean.
+Since we have categorical predictor variables, these priors allow the coefficients to adjust based on the data.
 
@@
 #| label: load-models
-library(rstanarm)
 poisson_model <-
@@
 mae_table |>
-  kable(col.names = c("Model", "Mean Absolute Error (MAE)")) 
+  kable(col.names = c("Model", "MAE")) |>
+  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
+
 ```
@@
 
+In @fig-mcmc, the MCMC algorithm is also used to check for potential issues with the model.
+The trace plot is constructed using a subset of categorical values from each predictor variable, as there are many values for each category.
+As seen in the trace plot in @fig-mcmc (a), the lines bounce around but remain horizontal, with a nice overlap between the chains.
+This indicates that the chains have effectively explored the posterior distribution [@textbook].
+The Rhat plot evaluates whether the chains have converged to a common distribution.
+As seen in the Rhat plot in @fig-mcmc (b), the values are close to 1 and fall below 1.1, suggesting no problems with convergence [@textbook].
+Therefore, the model appears to be properly converged, and we do not need to remove or modify predictors, adjust the priors, or re-run the model.
+
 # Results
 
-Our results are summarized in @tbl-model-results.
+To determine the reference levels for the categorical data outbreak settings, causative agent, and month, the model automatically selects the first category of each factor based on alphabetical orders of the category names.
+The intercept represents that the baseline outbreak duration is 2.257 days when all predictors are at their reference levels.
+@fig-90-interval shows that the
+
+## Outbreak Setting
+Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:
+- Hospital-Chronic Care (0.047): Outbreaks in this setting last 0.047 days longer than Hospital-Acute Care.
+- Hospital-Psychiatric (-0.080): Outbreaks in psychiatric hospitals last 0.080 days shorter than Hospital-Acute Care.
+- LTCH (Long-Term Care Homes) (0.465): Outbreaks in LTCHs last 0.465 days longer than in Hospital-Acute Care.
+- Retirement Home (0.343): Outbreaks in retirement homes last 0.343 days longer than in Hospital-Acute Care.
+
+## Causative Agent
+Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:
+
+Num. Obs. (3416): Based on 3416 observations.
+Algorithm (sampling): The model used Bayesian sampling for estimation.
 
 ```{r, message = FALSE, echo = FALSE}
-#| label: tbl-model-results
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| label: "fig-90-interval"
+#| fig-cap: "90 percent credibility interval for coefficients"
+tidy_neg_binomial <- tidy(neg_binomial_model, conf.int = TRUE, conf.level = 0.9)
 
-# modelsummary::modelsummary(
-#     list(
-#         "Model" = model
-#     ),
-#     statistic = "mad",
-#     fmt = 2
-# )
+ggplot(tidy_neg_binomial, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
+  geom_pointrange() +
+  coord_flip() +  
+  labs(
+    x = "Predictor",
+    y = "Coefficient Estimate"
+  ) +
+  theme_minimal() +
+  theme(
+    axis.text = element_text(size = 8),
+  )
 ```
@@
 
-In @fig-loo-comparison, we compare LOO performance of the Poission model against the negative binomial model based on the expected log pointwise predictive density (ELPD) and find that the negative binomial model has a higher ELPD value.
+In @tbl-loo-comparison, we compare LOO performance of the Poission model against the negative binomial model based on the expected log pointwise predictive density (ELPD) and find that the negative binomial model has a higher ELPD value.
 
@@
 
+## Model summary
+
+@tbl-model-summary presents a summary of the model used in the analysis, which includes the intercept and the coefficients for predictor variables, and the model fitting process.
+
+```{r, message = FALSE, echo = FALSE}
+#| label: tbl-model-summary
+#| tbl-cap: "Explanatory model of outbreak duration for Model (1), the negative binomial model"
+tidied_neg = tidy(neg_binomial_model)
+glance_neg = glance(neg_binomial_model)
+
+msummary_out = list(tidy = tidied_neg,
+                    glance = glance_neg)
+
+class(msummary_out) = 'modelsummary_list'
+
+modelsummary(
+    msummary_out, output = "kableExtra"
+) |>
+  kable_styling(font_size = 8)
+```
+
 ## Diagnostics
 
-<!-- ```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
+@fig-mcmc presents the diagnostic plots for the MCMC algorithm used to estimate the parameters of our model.
+These plots are essential for assessing the convergence of the sampling process and ensuring the reliability of the Bayesian estimates.
+
+```{r, message = FALSE, echo = FALSE}
+#| label: fig-mcmc
 #| fig-cap: "Checking the convergence of the MCMC algorithm"
@@
 #| layout-ncol: 2
+plot(neg_binomial_model, "trace", pars = c("(Intercept)", "outbreak_settingLTCH", "outbreak_settingRetirement Home", "causative_agentParainfluenza", "causative_agentNorovirus", "causative_agentEnterovirus/Rhinovirus", "monthJan", "monthJul", "monthDec"))
 
-plot(model, "trace")
-
-plot(model, "rhat")
-``` -->
+plot(neg_binomial_model, "rhat")
+```
 

### 2024-11-28 scripts/00-simulate_data.py

@@ -11,12 +11,12 @@
 import pandas as pd
 import numpy as np
 
-num_data = 500
+sample_size = 1000
 
 # Set random seed for reproducibility
-np.random.seed(853)
+np.random.seed(22)
 
-#### Simulate data ####
+#### Define Data Categories ####
 # Outbreak settings
 outbreak_settings = [
     "Hospital-Chronic Care",
@@ -37,29 +37,31 @@ causative_agent_groups = [
     "Respiratory syncytial virus",
 ]
 
-# Generate the data
+#### Simulate data ####
 outbreak_settings_sampled = np.random.choice(
     outbreak_settings,
-    size=num_data,
+    size=sample_size,
     replace=True,
 )
+
 causative_agents_sampled = np.random.choice(
     causative_agent_groups,
-    size=num_data,
+    size=sample_size,
     replace=True,
 )
+
 months_sampled = np.random.choice(
     list(range(1, 13)),
-    size=num_data,
+    size=sample_size,
     replace=True,
 )
-durations_sampled = np.random.randint(
-    3,
-    21,
-    size=num_data,
-)  # Random duration between 3 and 20 days
 
-# Create a polars DataFrame
+# Simulate durations using a normal distribution
+mean_duration = 14  
+std_duration = 14   
+durations_sampled = np.random.normal(loc=mean_duration, scale=std_duration, size=sample_size)
+
+# Create a Pandas DataFrame
 outbreak_data = pd.DataFrame(
     {
         "outbreak_setting": outbreak_settings_sampled,
@@ -68,7 +70,6 @@ outbreak_data = pd.DataFrame(
         "duration": durations_sampled,
     }
 )
-
 
 #### Save data ####
 # Write the simulated data to a CSV file

### 2024-11-28 scripts/03-clean_data.py

@@ -138,6 +138,23 @@ df = df[df["causative_agent"] != "Other"]
 rows_after = df.shape[0]
 rows_removed = rows_before - rows_after
 
+# Map month number to text
+month_mapping = {
+    1: "Jan",
+    2: "Feb",
+    3: "Mar",
+    4: "Apr",
+    5: "May",
+    6: "Jun",
+    7: "Jul",
+    8: "Aug",
+    9: "Sep",
+    10: "Oct",
+    11: "Nov",
+    12: "Dec",
+}
+df["month"] = df["month"].map(month_mapping)
+
 df = df[
     [
         "outbreak_setting",

### 2024-11-28 scripts/04-test_analysis_data.py

@@ -23,7 +23,7 @@ def test_column_count():
 def test_column_types():
     assert df["outbreak_setting"].dtype == "object", "'outbreak_setting' is not string"
     assert df["causative_agent"].dtype == "object", "'causative_agent' is not string"
-    assert df["month"].dtype == "int64", "'month' is not integer"
+    assert df["month"].dtype == "object", "'month' is not string"
     assert df["duration"].dtype == "int64", "'duration' is not integer"
 
 
@@ -39,15 +39,23 @@ def test_unique_outbreak_setting():
     ), "'outbreak_setting' should have at least 1 unique value"
 
 
-def test_month_range():
-    # Test that 'month' is in the valid range 1-12.
-    assert df["month"].between(1, 12).all(), "'month' contains invalid values"
+months = {"Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"}
 
+def test_month_values():
+    # Test that 'month' contains only valid month names.
+    
+    invalid_months = set(df["month"].unique()) - months # Set difference
+    assert not invalid_months, f"Dataset contains invalid months: {invalid_months}"
 
+def test_all_months_present():
+    # Test that all 12 months (Jan through Dec) are present in the 'month' column.
+    months_present = set(df["month"].unique())
+    missing_months = months - months_present # Set difference
+    assert not missing_months, f"Dataset is missing months: {missing_months}"
+    
 def test_duration_positive():
     # Test that 'duration' is positive.
     assert df["duration"].gt(0).all(), "'duration' contains non-positive values"
-
 
 def test_outbreak_setting_at_least_two_unique():
     # Test that 'outbreak_setting' contains at least two unique values.

### 2024-11-28 scripts/07-model_data.R

@@ -54,4 +54,4 @@ neg_binomial_model <- stan_glm(
 
 # Save the models for later use
 saveRDS(poisson_model, file = "models/poisson_model.rds")
-saveRDS(neg_binomial_model, file = "models/neg_binom_model.rds")
+saveRDS(neg_binomial_model, file = "models/neg_binomial_model.rds")

### 2024-11-28 scripts/08-check_models.R

@@ -29,18 +29,18 @@ test_data <- test_data |>
   )
 
 # # Posterior predictive checks
-# pp_check(poisson_model) +
-#   theme(legend.position = "bottom")
+pp_check(poisson_model) +
+  theme(legend.position = "bottom")
 
-# pp_check(neg_binomial_model) +
-#   theme(legend.position = "bottom")
+pp_check(neg_binomial_model) +
+  theme(legend.position = "bottom")
 
 # # Compare the models using LOO (Leave-One-Out Cross Validation)
-# poisson_loo <- loo(poisson_model, cores = 2)
-# neg_binomial_loo <- loo(neg_binomial_model, cores = 2)
+poisson_loo <- loo(poisson_model, cores = 2)
+neg_binomial_loo <- loo(neg_binomial_model, cores = 2)
 
-# loo_comparison <- loo_compare(poisson_loo, neg_binomial_loo)
-# loo_comparison
+loo_comparison <- loo_compare(poisson_loo, neg_binomial_loo)
+loo_comparison
 
 # Make predictions on the test data
 poisson_pred <- predict(poisson_model, newdata = test_data, type = "response")

### 2024-11-29 other/datasheet/datasheet_template.qmd
finalize paper and complete datasheet

---
title: "Datasheet for 'Outbreaks in Toronto Healthcare Institutions'"
author: 
  - Kevin Cai
thanks: "Code and data are available at: [https://github.com/kevicai/toronto-healthcare-outbreak-prediction](https://github.com/kevicai/toronto-healthcare-outbreak-prediction."
date: today
date-format: "D MMMM YYYY"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to track and manage outbreaks of gastroenteric and respiratory infections in Toronto healthcare institutions. It addresses the need for real-time, structured data on institutional outbreaks to support public health responses and decision-making.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The dataset was created by Toronto Public Health.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - Toronto government funded the of the dataset.
4. *Any other comments?*
    - No.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Each instance represents a recording of an outbreak event in a Toronto healthcare facility. The dataset includes the institution name and location, the type of outbreak, the causative agents (primary and secondary), and the start and end time of the outbreak.
2. *How many instances are there in total (of each type, if appropriate)?*
	- There are 5387 instances in total.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- The dataset contains all reported outbreaks from Toronto healthcare institutions.
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Each instance consists of features such as the institution name, address, outbreak setting, type of outbreak, causative agents, and the dates of outbreak onset and conclusion.
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- Each instance has a label that indicates whether the outbreak is active or not.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- The dataset includes instances where the second causative agent is "None" if only one causative agent is identified. 
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- There are no explicit relationships between instances, as each outbreak is treated as an independent observation.
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- No.
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- No.
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The dataset relies on external sources, specifically Toronto Public Health’s outbreak reporting system, which is publicly accessible. The data is guaranteed to be updated weekly, and archival versions are not specified.
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- No.
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No.
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- No.
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No.
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No.
16. *Any other comments?*
	- No.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- Data was acquired through mandatory reporting by healthcare institutions to Toronto Public Health, where outbreaks are monitored and tracked.
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Data was collected via public health reporting systems and APIs from the City of Toronto Open Data Portal, and verified through public health monitoring and oversight.
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- N/A.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- The data collection was conducted by Toronto Public Health staff and healthcare workers at the outbreak institutions, but their compensation is unknown.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- Data was collected continuously from January 2016 to the present and is updated weekly.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- No
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The data was obtained from Toronto Public Health and the City of Toronto Open Data Portal.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- The dataset is publicly available through the City of Toronto Open Data Portal. Specific notifications to individuals about data collection were not required as the data is aggregated and anonymized.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Consent was not required due to the nature of the dataset.
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Consent was not required due to the nature of the dataset.
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- No.
12. *Any other comments?*
	- No.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- Yes, the data has been cleaned and preprocessed. Instances with missing or irrelevant data were removed, and categorical variables were renamed for clarity. Additionally, dates were standardized, and the duration of outbreaks was calculated based on the start and end dates.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- The raw data is available from the City of Toronto Open Data Portal. Link to the raw dataset on the City of Toronto Open Data Portal.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- The preprocessing was done using Python and the pandas package. The code used for cleaning and analysis is available at GitHub repository.
4. *Any other comments?*
	- No.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- Yes, the dataset has been used for analysis in outbreak prediction modeling, as part of research on factors influencing the duration of outbreaks in healthcare institutions.
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- The code and data used for the analysis are available at GitHub repository.
3. *What (other) tasks could the dataset be used for?*
	- The dataset could be used for further research on outbreak prediction, healthcare preparedness, and policy recommendations. It could also be used to explore the impact of different infectious agents on outbreak duration or for developing risk models for healthcare settings.
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- The dataset contains aggregated information on outbreaks and does not include any personally identifiable data, so there are minimal risks regarding unfair treatment or harms. However, consumers of the dataset should be mindful of the limitations in the dataset, such as missing values or the inability to link data to individual patients, which could affect the accuracy of certain analyses.
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- The dataset should not be used for any analysis that requires personally identifiable information, as the data is anonymized and aggregated. It should not be used to draw conclusions about specific individuals or make personal health-related decisions.
6. *Any other comments?*
	- No.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- Yes, the dataset is publicly available through the City of Toronto Open Data Portal and can be accessed by third parties.
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- The dataset is available through the City of Toronto Open Data Portal as a downloadable CSV file. It does not currently have a DOI but is available via the public open data platform.
3. *When will the dataset be distributed?*
	- The dataset is refreshed weekly and distributed via the City of Toronto Open Data Portal.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- The dataset is distributed under the Open Government License - Toronto, which allows users to copy, modify, and distribute the data for any lawful purpose. Link to license details.
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No third-party IP-based restrictions apply to the dataset.
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- No.
7. *Any other comments?*
	- No.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- The dataset will be maintained by Toronto Public Health and updated weekly via the City of Toronto Open Data Portal.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- The contact email for Toronto Public Health is edau@toronto.ca.
3. *Is there an erratum? If so, please provide a link or other access point.*
	- No.
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- The dataset is updated weekly by Toronto Public Health. Consumers can track updates through the City of Toronto Open Data Portal.
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- The dataset does not relate to people.
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Older versions of the dataset are archived and available for reference through the City of Toronto Open Data Portal.
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- The dataset is open and available for public use through the City of Toronto Open Data Portal. Individuals wishing to contribute to or augment the dataset would need to work through Toronto Public Health or the City of Toronto’s data governance procedures.
8. *Any other comments?*
	- No.

\newpage

# References

### 2024-11-29 paper/paper.qmd
finalize paper and complete datasheet

@@
 title: "Analyzing the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
-subtitle: "My subtitle if needed"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This study investigates the factors influencing the duration of infectious disease outbreaks in healthcare institutions in Toronto, focusing on the role of outbreak setting, causative agent, and seasonal variations.
+Using a Bayesian negative binomial regression model, the analysis identifies significant drivers of outbreak duration and highlights the importance of targeted preparedness in settings like Long-Term Care Homes (LTCHs) and Retirement Homes.
+The findings suggest that COVID-19 has a substantial impact on outbreak duration compared to other respiratory viruses.
+This research underscores the need for robust healthcare infrastructure and adaptive strategies to manage future healthcare outbreaks effectively."
 format: pdf
@@
 
-Outbreaks of infectious diseases in healthcare institutions are a significant challenge, particularly in hospitals, long-term care homes, and retirement homes, where vulnerable populations are most at risk.
-Understanding the factors that influence the duration of these outbreaks is important for improving preparedness, response times, and resource allocation.
-Previous studies have looked at factors affecting outbreak durations, but more research is needed, especially in Toronto’s healthcare system.
-This paper analyzes the key factors that influence the duration of outbreaks in Toronto healthcare institutions, focusing on outbreak setting, causative agents, and the timing of outbreaks.
+Infectious disease outbreaks in healthcare settings present significant challenges, particularly in hospitals, long-term care homes, and retirement homes, where vulnerable populations are at highest risk [@torontoActiveOutbreaks].
+These environments house individuals who are more susceptible to infections, including the elderly, individuals with compromised immune systems, and those with underlying health conditions.
+In Toronto, public health authorities are required to respond to and manage outbreaks of gastroenteric and respiratory illnesses in these institutions [@torontoActiveOutbreaks].
+Efficient management of such outbreaks is essential to prevent widespread illness and ensure the safety of patients, staff, and visitors.
+Understanding the factors that influence the duration of these outbreaks is important for enhancing preparedness, minimizing response times, and optimizing resource allocation.
 
-This study uses a Bayesian negative binomial regression model to predict outbreak durations in Toronto's healthcare facilities.
-The model includes factors like the type of healthcare institution (outbreak setting), the causative agent, and the month the outbreak occurred.
-This research aims to identify which factors contribute most to variation in outbreak durations.
-It provides actionable information for public health officials and healthcare managers to improve their response strategies.
+The factors influencing outbreak duration can vary significantly depending on the healthcare setting, the causative agent of the illness, and the timing of the outbreak.
+In nursing homes, for example, the close living conditions, shared spaces, and frequent interactions between residents, staff, and visitors contribute to the spread of infections.
+In hospitals, the high turnover of patients and healthcare workers, coupled with the need for intensive medical care, can also worsen the problem.
+Additionally, the characteristics of specific pathogens can influence transmission dynamics and outbreak duration.
+For instance, the concentration of SARS-CoV-2 in patient secretions peaks approximately 10 days after symptom onset, influencing transmission patterns and the length of outbreaks in healthcare facilities [@sarsOnHealthcare].
+Understanding why some outbreaks persist longer than others requires identifying the drivers of these durations.
+While previous studies have examined factors affecting outbreak durations, more research is needed, especially within the context of Toronto's healthcare system.
+Toronto, with its large and diverse population, faces unique challenges in managing healthcare-associated outbreaks.
+This study aims to estimate the factors influencing the duration of infectious disease outbreaks in Toronto healthcare facilities, by examining various factors such as the type of healthcare setting, the causative agent, and the month when the outbreak occurred,
 
-The results of this analysis help manage outbreaks, optimize patient care, and reduce the strain on healthcare facilities.
-Identifying high-risk settings or times of year when outbreaks last longer can improve resource allocation and response planning.
-This study also adds to the literature on healthcare outbreaks, with a focus on Toronto’s healthcare system.
+To better understand the relationships between these factors and outbreak duration, this study employs a Bayesian negative binomial regression model.
+This statistical model is well-suited to handling count data and predicting the number of days an outbreak lasts.
+By identifying the drivers of outbreak duration, the study will help inform strategies for improving outbreak preparedness and response.
+These findings are important for optimizing resource allocation, ensuring that healthcare facilities are well equipped to handle outbreaks, and improving response efforts.
 
@@
 
-This report uses the Outbreaks in Toronto Healthcare Institutions dataset, contains data from January 2016 to November 2024.
+To examine the effects of factors in healthcare facilities in Toronto on outbreak duration, this report uses the Outbreaks in Toronto Healthcare Institutions dataset, which contains data from January 2016 to November 2024.
 The dataset is provided by Toronto Public Health, through City of Toronto Open Data Portal [@datasource].
+Since no other datasets specific to healthcare institutions in Toronto are available, no additional data sources were considered.
 The dataset tracks reported outbreaks of gastroenteric and respiratory illnesses in Toronto healthcare institutions and contains detailed information on outbreak settings, causative agents, and outbreak durations.
@@
 
-### Outbreak Setting
+### Outbreak setting
 The Outbreak Setting variable is categorical and identifies the type of healthcare institution where the outbreak occurred, such as hospitals, long-term care homes (LTCH), or retirement homes.
@@
 
-### Causative Agent
+### Causative agent
 The Causative Agent variable is categorical and reflects the infectious agents responsible for outbreaks.
@@
 
-### Month Outbreak Began
+### Month outbreak began
 The Month variable is categorical and records the calendar month when each outbreak started as a name (e.g., Jan, Feb).
@@
 
-# Model
+# Model {#sec-model}
 
-## Model Overview
+## Model overview
 
@@
 
-## Model Setup
+## Model setup
 
@@
 
-## Model Selection
+## Model selection
 
@@
 
-## Model Diagnostics and Validation
+## Model diagnostics and validation
 
@@
 
-# Results
+# Results {#sec-results}
 
-To determine the reference levels for the categorical data outbreak settings, causative agent, and month, the model automatically selects the first category of each factor based on alphabetical orders of the category names.
+Our results are summarized in @tbl-model-summary using the modelsummary package [@modelsummary], where the model is tidied with the broom.mixed package [@broom] and the output table is scaled down with the kableExtra package [@kableExtra].
+In the summary the model automatically selects the first category of each factor based on alphabetical orders of the category names.
 The intercept represents that the baseline outbreak duration is 2.257 days when all predictors are at their reference levels.
-@fig-90-interval shows that the
-
-## Outbreak Setting
-Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:
-- Hospital-Chronic Care (0.047): Outbreaks in this setting last 0.047 days longer than Hospital-Acute Care.
-- Hospital-Psychiatric (-0.080): Outbreaks in psychiatric hospitals last 0.080 days shorter than Hospital-Acute Care.
-- LTCH (Long-Term Care Homes) (0.465): Outbreaks in LTCHs last 0.465 days longer than in Hospital-Acute Care.
-- Retirement Home (0.343): Outbreaks in retirement homes last 0.343 days longer than in Hospital-Acute Care.
-
-## Causative Agent
-Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:
-
-Num. Obs. (3416): Based on 3416 observations.
-Algorithm (sampling): The model used Bayesian sampling for estimation.
+For the outbreak setting, the reference category is "Hospital-Acute Care", all other outbreak settings are compared to "Hospital-Acute Care" to evaluate their influence on outbreak duration.
+The reference category is "Coronavirus" for the causative agent, the reference category is "April" for the month.
+By comparing the results, we can observe how each variable influences the duration differently.
 
@@
 
-# Discussion
+Based on the 90% credibility interval plot for the model in @fig-90-interval, we can analyze the influence of each predictor on outbreak duration:
+Intercept: The intercept coefficient estimate lies above 0, indicating that the baseline outbreak duration (when all predictors are at their reference levels) is positively associated with longer outbreak durations.
+The confidence intervals for most variables are relatively small, with the largest interval reaching around 0.25.
+A wider interval suggests greater uncertainty and higher variation in the estimate for that predictor.
+Conversely, narrower intervals indicate more precise estimates of the effect of the predictor on outbreak duration.
 
-## First discussion point {#sec-first-point}
+Based on @fig-90-interval and @tbl-model-summary, we can see that for outbreak settings, the model suggests that outbreaks in Retirement Homes and Long-Term Care Homes (LTCHs) tend to last longer than other settings.
+This supports the expectation that these facilities, housing more vulnerable demographics, are more prone to prolonged outbreaks.
+On the other hand, Hospital-Psychiatric outbreaks tend to be shorter, as indicated by the negative coefficient.
+This supports the variability in duration across different healthcare settings.
+Regarding the causative agents, Respiratory Syncytial Virus (RSV), Parainfluenza, Norovirus, Enterovirus/Rhinovirus, and COVID-19, all show positive coefficient estimates, with COVID-19 being much larger compared to all other agents.
+This suggests that outbreaks caused by these agents last longer than those caused by Coronavirus (the reference group).
+In contrast, Influenza outbreaks are expected to last shorter, as indicated by its negative coefficient, while Metapneumovirus shows a marginal increase in duration, though its coefficient is very close to 0.
+Finally, the month of the outbreak also plays a role in its duration.
+The model suggests that outbreaks occurring in August and January are generally shorter compared to April (the reference month), while December outbreaks tend to last longer, which may suggest seasonal variation in outbreak duration.
+Overall, the 90% credibility intervals for the coefficients highlight the significance of the predictors in influencing outbreak duration, with some factors such as the type of healthcare setting and causative agent having a more substantial effect, while others, like month of occurrence, show moderate influences.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+# Discussion {#sec-discussion}
 
-## Second discussion point
+## Long-term care homes and retirement homes are most susceptible to outbreaks
+The type of healthcare setting where an outbreak occurs significantly impacts its duration.
+According to the model’s results, outbreaks in Long-Term Care Homes (LTCHs) and Retirement Homes are more likely to last longer compared to other settings.
+This finding aligns with the vulnerability of populations in these settings, particularly the elderly, who may have weakened immune systems or multiple preexisting conditions.
+This observation is consistent with a population-based study of Retirement Homes (RHs) in Ontario, which reported that RHs are highly susceptible to SARS-CoV-2 outbreaks due to home-level and community-level factors.
+Larger resident capacities, co-location with Long-Term Care facilities, and offering multiple services onsite were identified as significant contributors to outbreak risk.
+These factors suggest increased staff movement and staff-resident interactions may exacerbate outbreak dynamics in these settings [@retirementCovidRisks].
+Furthermore, the study observed that RHs in regions with higher community-level ethnic concentration or increased regional COVID-19 incidence were at elevated risk of outbreaks, although the sensitivity to these factors diminished in Wave 2 due to improved preventative measures, such as mandatory surveillance testing and stricter infection control practices [@retirementCovidRisks].
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+The facilities of Long-Term Care Homes and Retirement Homes often lack the resources and infrastructure of hospital settings, which may contribute to longer outbreak durations in these environments.
+Previous research highlights staffing shortages in Long-Term Care Homes further worsen these challenges.
+The Long-Term Care workforce, has faced retention and recruitment issues for years.
+The COVID-19 pandemic intensified these problems, leading to burnout and workforce attrition despite wage increases and hiring incentives [@longTermCareStaffing].
+Additionally, limited training and career advancement opportunities for direct care aides further strain care quality.
+These staffing issues, combined with high operational costs, hinder the ability of LTCH facilities to manage outbreaks effectively [@longTermCareStaffing].
+Addressing these systemic challenges is critical to improving outbreak management in LTCH and Retirement Homes.
 
-## Third discussion point
+## The Impact of COVID-19 vs regular Coronavirus
 
-## Weaknesses and next steps
+The analysis shows that COVID-19 had a significant impact on the duration of outbreaks in healthcare settings, especially when compared to other respiratory pathogens like the common coronavirus.
+COVID-19 outbreaks were found to last much longer than those caused by regular coronaviruses, a finding consistent with the global experience during the pandemic.
+Several factors contributed to this prolonged duration, including the high transmissibility of the virus, its asymptomatic spread, and the challenges of managing outbreaks in healthcare settings housing vulnerable populations [@retirementCovidRisks].
+Additionally, the virus’s evolving nature, with new variants emerging over time, further complicated efforts to contain it, requiring healthcare facilities to frequently adjust infection control measures.
 
-Weaknesses and next steps should also be included.
+While the model suggests that regular coronaviruses did not have a similar effect on outbreak duration, as seen in @tbl-model-summary, the coefficient is close to that of other causative agents.
+This indicates that once the COVID-19 pandemic is under control, coronaviruses may not pose the same level of challenge to healthcare systems.
+The findings highlight the importance of ongoing preparedness for future global health crises, where the emergence of new infectious diseases or highly transmissible variants could once again stretch healthcare resources.
+Continued investment in healthcare infrastructure, surveillance, and rapid response capabilities will be essential to minimize the potential impact of future pandemics on healthcare facilities, ensuring they are well-equipped to manage outbreaks of varying durations.
 
-\newpage
+## The Effect of Winter on Outbreak Duration
+The model suggests that outbreaks occurring in winter, particularly in December, tend to last longer compared to other months.
+This seasonal variation is likely due to higher rates of respiratory illnesses such as influenza and RSV during colder months, combined with holiday-related gatherings and staff shortages.
+The strain on healthcare facilities, particularly those housing vulnerable populations, can further extend outbreak durations.
+These findings highlight the need for targeted preparedness during winter, including increased staffing, early vaccinations, and robust infection control measures to minimize the impact of seasonal outbreaks.
 
-\appendix
+## Weaknesses and Next Steps
+
+One limitation of this analysis is the reliance on available outbreak data, which may not fully capture all outbreaks in healthcare settings.
+For instance, some incidents might have gone unreported, or data may have been misclassified, which could introduce bias or reduce the generalizability of the findings.
+Additionally, the model does not account for interactions between different factors, such as the combined effects of outbreak setting and causative agent, which could provide more nuanced insights.
+The assumptions made in the negative binomial regression model also rely on the availability of complete and accurate data, which may not always be the case, particularly with retrospective datasets.
+Furthermore, although the Mean Absolute Error (MAE) of 6.52 days is relatively small, it still represents a notable deviation from the actual outbreak duration, suggesting that further investigation and model refinement are needed to improve accuracy.
+
+Future research could enhance this study by incorporating more detailed data, including healthcare interventions and patient information.
+Exploring the interaction between predictors like outbreak setting, causative agent, and seasonal trends could provide deeper insights into outbreak dynamics.
+The model also does not incorporate the potential influence of COVID-19 vaccinations on outbreak duration, an important factor that could have significantly altered the course of outbreaks, particularly in healthcare settings.
+Further research could explore the role of vaccination coverage and its interaction with other variables to improve predictive accuracy.
 
@@
 
-# Additional data details
+## Model details
 
-# Model details {#sec-model-details}
-
-## Outcome Variable Variance and Mean
-
-## Posterior predictive check
+### Posterior predictive check
 
@@
 
-## Leave-One-Out (LOO) Cross Validation (CV) Comparison
+### Leave-One-Out (LOO) Cross Validation (CV) Comparison
 
@@
 
-## Model summary
+### Model summary
 
@@
 
-## Diagnostics
+### Diagnostics
 

### 2024-11-29 scripts/00-simulate_data.py

@@ -35,6 +35,7 @@ causative_agent_groups = [
     "Norovirus",
     "Parainfluenza",
     "Respiratory syncytial virus",
+    "Coronavirus",
 ]
 
 #### Simulate data ####

### 2024-11-29 scripts/05-exploratory_data_analysis.py

@@ -15,26 +15,7 @@ script_dir = os.path.dirname(os.path.abspath(__file__))
 df = pd.read_csv(os.path.join(script_dir, "../data/02-analysis_data/analysis_data.csv"))
 
 # See unique settings
-unique_values = sorted(df["outbreak_setting"].unique())
-print(len(unique_values))
-for value in unique_values:
-    print(value)
-
-# See unique causative agents
 unique_values = sorted(df["causative_agent"].unique())  # original 55 agents
 print(len(unique_values))
 for value in unique_values:
     print(value)
-
-# Top 10 agents
-top_10 = (
-    df["causative_agent"]
-    .value_counts()
-    .sort_values()  # Sort by counts in descending order
-    .head(10)
-)
-print(top_10)
-
-# Count causative agents occurances
-value_counts = df["causative_agent"].value_counts()
-print(value_counts)

### 2024-12-02 paper/paper.qmd
add graph sketches and idealized survey

@@
 format: pdf
+toc: true
 number-sections: true
@@
     y = "Duration (days)",
-    title = "Duration of Outbreaks Across Months"
   ) +
@@
     axis.title = element_text(size = 10),
-    plot.title = element_text(size = 12, face = "bold")
   )
@@
 
-# Appendix {-}
+# Appendix {sec-appendix}
 
@@
 
+## Idealized Survey
+### Introduction
+Thank you for participating in this survey!
+We are collecting data from healthcare workers to better understand infectious disease outbreaks in healthcare institutions.
+Your responses will help inform strategies for improving outbreak management and preparedness.
+
+The survey is expected to take approximately 10 minutes to complete. All responses will be kept confidential and used solely for research purposes.
+
+If you have any questions, please contact:
+
+- **Survey Coordinator**: Kevin Cai  
+- **Email**: [kev.cai@mail.utoronto.ca](mailto:kev.cai@mail.utoronto.ca)
+
+### Survey Questions
+
+#### Section 1: Institutional Information
+
+1. **Institution Name**  
+   Open-ended response:  
+   \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+2. **Institution Address**  
+   Open-ended response:  
+   \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+3. **Outbreak Setting** (Select one):  
+   - Hospital - Acute Care  
+   - Hospital - Chronic Care  
+   - Hospital - Psychiatric  
+   - Long-Term Care Home (LTCH)  
+   - Retirement Home  
+   - Other: \_\_\_\_\_\_\_\_\_
+
+#### Section 2: Outbreak Information
+
+4. **Type of Outbreak** (Select one):  
+   - Respiratory  
+   - Enteric  
+   - Other: \_\_\_\_\_\_\_\_\_
+
+5. **Primary Causative Agent** (Select one):  
+   - COVID-19  
+   - Influenza  
+   - Norovirus  
+   - Respiratory syncytial virus  
+   - Parainfluenza  
+   - Enterovirus/Rhinovirus  
+   - Metapneumovirus  
+   - Coronavirus (non-COVID)  
+   - Other: \_\_\_\_\_\_\_\_\_
+
+6. **Secondary Causative Agent (if applicable)** (Select one or leave blank):  
+   - COVID-19  
+   - Influenza  
+   - Norovirus  
+   - Respiratory syncytial virus  
+   - Parainfluenza  
+   - Enterovirus/Rhinovirus  
+   - Metapneumovirus  
+   - Coronavirus (non-COVID)  
+   - Other: \_\_\_\_\_\_\_\_\_
+
+#### Section 3: Timeline and Status
+
+7. **Date Outbreak Began**  
+   Date picker: \_\_\_\_\_\_\_\_\_
+
+8. **Date Outbreak Was Declared Over**  
+   Date picker: \_\_\_\_\_\_\_\_\_
+
+9. **Is the Outbreak Currently Active?** (Select one):  
+   - Yes  
+   - No
+
+#### Section 4: Outbreak Duration and Impact
+
+10. **Duration of Outbreak (in days)**  
+    Open-ended response:  
+    \_\_\_\_\_\_\_\_\_
+
+11. **Did the outbreak require external resources or assistance (e.g., public health interventions)?**  
+    - Yes  
+    - No
+
+12. **What were the primary challenges faced during this outbreak?** (Select all that apply):  
+    - Staffing shortages  
+    - Supply chain issues (e.g., PPE, medications)  
+    - Limited isolation facilities  
+    - Insufficient training or protocols  
+    - Other: \_\_\_\_\_\_\_\_\_
+
+13. **What mitigation measures were most effective in managing this outbreak?** (Select all that apply):  
+    - Enhanced cleaning and disinfection  
+    - Improved PPE availability  
+    - Isolation/quarantine measures  
+    - Staff education and training  
+    - Rapid diagnostic testing  
+    - Other: \_\_\_\_\_\_\_\_\_
+
+#### Section 5: Feedback and Additional Information
+
+14. **In your opinion, what could have been done differently to reduce the duration of this outbreak?**  
+    Open-ended response:  
+    \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+15. **Do you have additional comments or observations about this outbreak?**  
+    Open-ended response:  
+    \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
+### Thank You
+
+Thank you for completing this survey! Your responses will contribute to improving outbreak management strategies in healthcare institutions.
+
+If you would like to receive updates about this research or a copy of the final report, please provide your email address below (optional):  
+_Email Address:_  
+\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
+
 \newpage

### 2024-12-02 paper/paper.qmd
rephrase introduction

@@
 
-The factors influencing outbreak duration can vary significantly depending on the healthcare setting, the causative agent of the illness, and the timing of the outbreak.
+This study estimates the duration of infectious disease outbreaks in Toronto healthcare facilities.
+Specifically, it examines how factors such as the type of healthcare setting, the causative agent, and the month when the outbreak began influence the number of days an outbreak lasts.
+By analyzing these relationships, the study evaluates how changes in these predictors affect outbreak duration.
+
+The duration of an outbreak is influenced by various factors  including the healthcare setting, the causative agent, and the timing of the outbreak.
 In nursing homes, for example, the close living conditions, shared spaces, and frequent interactions between residents, staff, and visitors contribute to the spread of infections.
@@
 Toronto, with its large and diverse population, faces unique challenges in managing healthcare-associated outbreaks.
-This study aims to estimate the factors influencing the duration of infectious disease outbreaks in Toronto healthcare facilities, by examining various factors such as the type of healthcare setting, the causative agent, and the month when the outbreak occurred,
-
-To better understand the relationships between these factors and outbreak duration, this study employs a Bayesian negative binomial regression model.
-This statistical model is well-suited to handling count data and predicting the number of days an outbreak lasts.
-By identifying the drivers of outbreak duration, the study will help inform strategies for improving outbreak preparedness and response.
-These findings are important for optimizing resource allocation, ensuring that healthcare facilities are well equipped to handle outbreaks, and improving response efforts.
+To analyze these relationships, the study uses a Bayesian negative binomial regression model, suitable for count data and predicting the duration of outbreaks.
+The findings aim to inform strategies for better outbreak management, resource allocation, and preparedness in healthcare facilities.
 
